A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-03-07 analysis/paper/paper.Rmd
Set up plotting with HDR. Also miss a few sites for sites_sl due to a previous error in the subsetting code which caused a few sites to be excluded

The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g. @akerlund1996; @bjerck2005; @ilves2011; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009].

@@

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g. @hughes2016; @stroeven2016, see Figure \@ref(fig:deglaciation)], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g. @mörner1979; @svendsen1987]. In other words, the RSL has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable RSL or been subject to marine transgression [e.g. @romundset2015; @svendsen1987]. These conditions are directly reflected in the archaeological record. In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [@reitan2009]. Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008; @glorstad2020]. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @romundset2018; @sørensen1979]. This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [ @berg-hansen2009; @bergsvik2009]. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm is to have led site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2012; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2011a]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227]. @nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g. @amundsen2006; @solheim2012, 74; @ostmo1988]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example common in survey projects, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g. @berg-hansen2009; @eskeland2017]. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the typologically informed date-span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are few [see also @ilves2011], and are typically conducted on a site-by-site or project-by-project basis, meaning the assumed shore-bound site location is largely based on a mosaic of smaller scale investigations.

One of the more comprehensive evaluations of this relationship was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 ^14^C-dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites. However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates. This is based on typological and technological characteristics of the assemblages. Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate [e.g. @persson2008; @schulke2020]. However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas evidenced by artefact inventories or multiple site features, and not remnants of stays as ephemeral to only be discernible by individual features or dubious ^14^C-dates. The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.

@@

Shoreline dating could be termed a chorological dating method. That is, it is a method where the target phenomenon is given a likely date of occurence with reference to its geographical location -- in this case its altitude relative to the present day sea-level. The method is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and associated uncertainty to these dimensions. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location from the displacement curves using inverse distance weighting based on the distance between the site and the isobases of the displacement curves [e.g. @conolly2020; @conolly2006, 94--97]. This was done for each year along the entirety of the curves, weighting the interpolation by the squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same settlement phase. In the case where there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.

@@

Where x is [...] In Figure \@ref(fig:shoredate) this formula is used to shoreline date the same sites from where this relationship was derived. Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some deviance in elevation. The result of comparing the synchroneity of the dates using the method presented by @parnell2008 is given in Figure \@ref(fig:shoredate)B. Here, 100,000 age samples drawn from each shoreline date was subtracted from 100,000 age samples drawn from the corresponding ^14^C-date. The resulting range of the 95% highest density region (HDR, @ ) was then checked to see if it crosses zero, in which case the dates are considered synchronous. The result of this procedure indicates that the shoreline date correspond with the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%). The deviation from the shoreline date at Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langemyr and Kvastad A2. For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B), and in the simulation of the distance to the shoreline this was simulated to be 0m across virtually all of the 1000 simulation runs for both sites (see supplementary material). If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of the two Løsvås sites and Gunnarsrød 5 are instead due to extraneous factors and not the method itself, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated.

@@

```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory. The sites are dated using the mean elevation of the site polygon and by employing the exponential decay function identified in Figure 7 for characterising the distance between sites and shoreline. A) The 95\\% probability range for the shoreline dates are plotted in gray and underlined in red. These are plotted against the oldest radiocarbon-informed phase for each site in black. B) Using the method described by Parnell et al. (2008), the synchroneity of the dates is evaluated. When the 95\\% interval for age difference crosses zero the dates are considered synchronous.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

@@

In Figure \@ref(fig:redate), excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites are subjected to the method for shoreline dating outlined above. These are compared to the dates originally proposed in the excavation reports for the sites. To avoid issues with recent disturbances in the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site. This is an illustrative, but unfair exercise for a few reasons. First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit but undefined uncertainty range. Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change. Finally, they are often only meant to indicate a lower bound for when the sites could have been in use. I do, however, think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner. The results from applying the method devised here highlights the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% intervals for the dates. In some cases it provides a very precise date range and in others it offers little more than a *terminus post quem*. This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise date ranges (Figure \@ref(fig:map-iso)B).  However, a few of the date ranges extend well beyond major chronological divisions, even the entire Stone Age, and could be severely and securely constrained with only cursory reference to typology. This thus points to the possibility of drawing on other temporal data such as typological indicators or radiocarbon dates, for example within a Bayesian framework, to improve the precision of the dating of the sites.

```{r redate, echo = FALSE, fig.cap = "Shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories. The 95\\% probability ranges in red are comapred to the dates originally proposed by the excavation reports in black.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/redate.png"))
```

Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g. @bang-andersen2012; @breivik2014; @glorstad2016]. The shoreline dated pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a]. These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision use of the sites after the sea retreated any significant distance from their location. In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically. A comparison of his results and the ones achieved using the above-outlined approach are given in Figure \@ref(fig:brunlanes)A. The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here [cf. @sorenseninpress]. Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites were to be considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they don't overlap. This information is not integrated in the approach outlined here, but could justify reducing the the uncertainty ranges. Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does nonetheless result in very precise dates. Above it was shown how additional temporal data could be combined with the method to improve its accuracy and precision. This example, on the other hand, highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use. One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the steepness of the slope to the shoreline. Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline for when the site was in use. Such approaches would make less of an impact in this setting, where the 95% probability range is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change as been less severe (cf. Figure \@ref(fig:redate)).

```{r brunlanes, echo = FALSE, fig.cap = "Shoreline dating of the Brunlanes sites using site altitudes provided by Jaksland (2014:tab.4). A) The result of applying the approach to shoreline dating outlined above. Dates provided by Jaksland (2014) are plotted in black. The box indicates a 50 year uncertainty range which in combination with the black line extends 200 years. Follwing Jaksland's approach these start from the earliest possible date. B) Map showing the centroids of the Pauler sites and Sky 1. The sea-level has been simulated using the probability density associated with the shoreline date for Pauler 1 (see also map in Jaksland 2014:fig.12a). Pauler 1 is the red point.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/brunlanes.png"))
```

>>>>

The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g. @akerlund1996; @bjerck2005; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009; @wikell2009].

@@

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g. @hughes2016; @stroeven2016, see Figure \@ref(fig:deglaciation)], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g. @mörner1979; @svendsen1987]. In other words, the RSL has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable RSL or been subject to marine transgression [e.g. @romundset2015; @svendsen1987]. These conditions are directly reflected in the archaeological record. In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [e.g. @hagen1963; @reitan2009]. Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008; @glorstad2020]. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @romundset2018; @sørensen1979]. This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [discussed in @berg-hansen2009; @bergsvik2009]. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm is to have led site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2012; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2011a]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227]. @nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g. @amundsen2006; @solheim2012, 74; @ostmo1988]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example common in survey projects, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g. @berg-hansen2009; @eskeland2017]. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves it is common to assume a shore-bound site location, even if the typologically informed date-span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014; @johansen1963; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are few [see also @ilves2011], and are typically conducted on a site-by-site or project-by-project basis, meaning the assumed shore-bound site location is largely based on a mosaic of smaller scale investigations.

One of the more comprehensive evaluations of the relationship between radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found an overlap between the probability density of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites. However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates. This is based on typological and technological characteristics of the assemblages. Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these are entirely erroneous results of contamination can be difficult to evaluate [e.g. @persson2008; @schulke2020]. However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas evidenced by artefact inventories or multiple site features, and not remnants of stays as ephemeral to only be discernible by individual features or dubious ^14^C-dates. The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.

The method of shoreline dating has variably been said to contribute a method for relative dating for sites located at different elevations, for offering a lower limit for the site could have been in use, or been used to date a larger number of sites assign uniform uncertainty ranges of 200 or 500 years.

@@

The method of shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and associated uncertainty to these dimensions. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location from the displacement curves using inverse distance weighting based on the distance between the site and the isobases of the displacement curves [e.g. @conolly2020; @conolly2006, 94--97]. This was done for each year along the entirety of the curves, weighting the interpolation by the squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7% probability, meaning these were effectively taken to represent the same settlement phase. In the case where there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.

@@

Where x is [...] In Figure \@ref(fig:shoredate) this formula is used to shoreline date the same sites from where this relationship was derived. Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some deviance in elevation. The result of comparing the synchroneity of the dates using the method presented by @parnell2008 is given in Figure \@ref(fig:shoredate)B. Here, 100,000 age samples drawn from each shoreline date was subtracted from 100,000 age samples drawn from the corresponding ^14^C-date. The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered synchronous. The result of this procedure indicates that the shoreline date correspond with the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%). The deviation from the shoreline date at Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langemyr and Kvastad A2. For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B), and in the simulation of the distance to the shoreline this was simulated to be 0m across virtually all of the 1000 simulation runs for both sites (see supplementary material). If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of the two Løsvås sites and Gunnarsrød 5 are instead due to extraneous factors and not the method itself, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated.

@@

```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory. A) The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black. These are plotted against the oldest radiocarbon-informed phase for each site in red. B) Using the method described by Parnell et al. (2008), the synchroneity of the dates is evaluated. When the range of the 95\\% HDR for age difference crosses zero the dates are considered synchronous.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

@@

To further explore the implementation for shoreline dating outline above, excavated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the method outlined above (Figure \@ref(fig:redate)). These are compared to the dates originally proposed in the excavation reports for the sites. To avoid issues with recent disturbances in the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site. This is an illustrative, but unfair exercise for a few reasons. First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit but undefined uncertainty range. Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change. Finally, they are often only meant to indicate a lower bound for when the sites could have been in use. I do, however, think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner. The results from applying the method devised here highlights the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% intervals for the dates. In some cases it provides a very precise date range and in others it offers little more than a *terminus post quem*. This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise date ranges (Figure \@ref(fig:map-iso)B).  However, some of the date ranges extend well beyond major chronological divisions, even the entire Stone Age, and could be severely and securely constrained with only cursory reference to typology. This thus points to the possibility of drawing on other temporal data such as typological indicators or radiocarbon dates, for example within a Bayesian framework, to improve the precision of the dating of the sites.

```{r redate, echo = FALSE, fig.cap = "Shoreline dated sites in the study area without radiocarbon dates or with radiocarbon dates that do not correspond to the artefact inventories. The range of the 95\\% HDR in black are comapred to the dates originally proposed by the excavation reports in red.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/redate.png"))
```

Not least following from the fact that very few pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g. @bang-andersen2012; @breivik2014; @glorstad2016]. The shoreline dated pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a]. These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision use of the sites after the sea retreated any significant distance from their location. In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically. A comparison of his results and the ones achieved using the above-outlined approach are given in Figure \@ref(fig:brunlanes)A. The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here [cf. @sorenseninpress]. Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites were to be considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they don't overlap. This information is not integrated in the approach outlined here, but could justify reducing the the uncertainty ranges. Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does nonetheless result in very precise dates. Above it was shown how additional temporal data could be combined with the method to improve its accuracy and precision. This example, on the other hand, highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use. One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the steepness of the slope to the shoreline. Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline for when the site was in use. Such approaches would make less of an impact in this setting, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change as been less severe (cf. Figure \@ref(fig:redate)).

```{r brunlanes, echo = FALSE, fig.cap = "Shoreline dating of the Brunlanes sites using site altitudes provided by Jaksland (2014:tab.4). A) The result of applying the approach to shoreline dating outlined above. Dates provided by Jaksland (2014) are plotted in red. The box indicates a 50 year uncertainty range which in combination with the red line extends 200 years. Follwing Jaksland's approach these start from the earliest possible date. B) Map showing the centroids of the Pauler sites and Sky 1. The sea-level has been simulated using the probability density associated with the shoreline date for Pauler 1 (see also map in Jaksland 2014:fig.12a). Pauler 1 is the red point.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/brunlanes.png"))
```
